---
title: "Toronto is becoming more bigoted: An analysis of hate crime in Toronto over time and groups"
subtitle: "My subtitle if needed"
author: 
  - Justin Klip
thanks: "Code and data are available at: https://github.com/justinklip/hate_crime_toronto_paper."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
analysis_data <- read_csv(here::here("data/analysis_data/hate_crime_analysis_data.csv"))


```

Talk way more about it.

# Results


Our results are summarized in @tbl-modelresults.

```{r}

```

```{r}

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

# Limitations and Next Steps

Weaknesses and next steps should also be included.

\newpage

\appendix



# Appendix {.unnumbered}



```{r}

```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}

```

\newpage

# References
