---
title: "Toronto's Uptick in Hate Crime and It's Differing Victims Over Time"
subtitle: "An Exploratory Analysis of Hate Crime in Toronto"
author: 
  - Justin Klip
thanks: "Code and data are available at: https://github.com/justinklip/hate_crime_toronto_paper."
date: today
date-format: long
abstract: "Using data from Open Data Toronto, this paper the tracks the evolution of hate crime in Toronto from 2018 to the end of 2023. The up tick of hate crime began during the start of COVID-19, remained at elevated levels since then, and climbed in 2023 to reach 3 times the 2018 level. The most common types of hate crime during this time period were religion and race based hate crimes. It is also shown that the composition of hate crime types varied largely in this time period, suggesting different motivations for hate crime in different time periods." 
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tinytex)
library(zoo)
library(tidyr)
library(opendatatoronto)
library(knitr)
library(treemapify)
library(lintr)

#Data Acquisition
file_path <- here::here("data/analysis_data/hate_crime_analysis_data.csv")
hate_crime_analysis_data <- read_csv(file_path)
```

# Introduction {#sec-introduction}

In 1988 the Canadian Parliament passed the Canadian Multiculturalism Act, an act that served to recognize and preserve Canada's multicultural heritage @government_of_canada_about_2024. Toronto serves as a shining example of Canada's multiculturalism, with over half of the population of Toronto being foreign-born [@mcneil_how_2023]. Yet, despite Toronto's reputation as an international city with large widespread cultural heterogeneity, hate crimes have seen an uptick in recent times. Hate crimes in Toronto have increased nearly 55% in 2024 in comparison to 2023 [@noauthor_hate_2024]. A hate crime is defined as a "criminal act done by a person who is motivated by an extreme bias or hatred towards a particular social group", and can vary all the way from vandalism to murder [@government_of_canada_victims_2023]. Acts like these are fundamentally at odds with the sentiment expressed by the Canadian Multiculturalism Act, and raise questions about Toronto's Toronto's multicultural status [@perry_disrupting_2015]. Although hate crime on a broad level has been studied before, an exploratory study on the dynamics of hate crime in Toronto can contribute to the literature by informing other researchers of trends in a unique population that is much more diverse than other places.

In this paper, I explore the dynamics of Toronto's Hate Crimes from 2018 to 2023, before the current uptick. In particular, I describe how the types of hate crimes, the types of victims, and the quantity of victims have changed over time. I find that that hate crimes have grown significantly since 2018 due to large increases in both 2020 and 2023, with more than 300 in comparison to just over 100 in 2018. Post 2020 the number of hate crimes also did not increase and instead remained high relative to their pre 2020 level, showing COVID-19 was not just a temporary shock to hate crimes. I find that the most common victims of hate crimes are Black or Jewish, and that religious groups and race groups are the most common groups to fall victim of hate crime. Lastly, I find that religion based hate crimes are a large portion of the 2023 rise in hate crimes in Toronto. These findings are all provide important information to policy makers in order to know which groups in Toronto are most vulnerable.

The rest of the paper is split into two main parts: @sec-data and @sec-discussion-and-limitations. @sec-data has multiple sub parts, @sec-measurement describes how the data was collected and measured from real world events. @sec-cleaning-and-packages describes how the data was cleaned and what packages were used in the analysis. @sec-observations  displays what a typical data point looks like by taking a random sample. @sec-data-and-variable-selection provides an explanation for the specific variable focus of the paper. @sec-data-analysis-and-plots plots the data in numerous ways to demonstrate trends of hate crime in Toronto. The second section, @sec-discussion-and-limitations broadly generalizes the results, what they mean, and what should be done with them. It also describes what this paper cannot do and points out room in the literature for expansion on this topic.

```{r}

```

# Data {#sec-data}

## Measurement {#sec-measurement}

To run my analysis, data was collected from Open Data Toronto's Open Data Portal, through the R library opendatatoronto [@opendatatoronto]. The data source used was their Hate Crimes Open Data data set [@noauthor_open_nodate] . This data of about 1400 observations has every crime that has been classified as a 'hate crime' by Toronto Police from the beginning of 2018 to the end of 2023. The data is generated as follows: a crime is committed in Toronto, and generally within a month that crime is reported by the victim. If a hate-motivated offence is suspected, the investigation is then led by a divisional investigator of the Hate Crime Unit (HCU) or by the HCU itself (Toronto Police Service). If hate-motivated offences are found, then this case will be added next year to the data set. Details of the crime are documented –- including what social group they were biased against –- these bias variables include ethnicity, race, gender, age, language, religion, and sexual orientation. Other variables such as date and time (of both the crime and the reporting of it), offence committed, neighborhood, and place (such as apartments, malls, parks) were also recorded. Every unique case number was also attributed to each observation in the data set. There were almost no empty observations, except for a few cases where place data was missing. @sec-cleaning-and-packages discusses this missing data further.

## Cleaning and Packages {#sec-cleaning-and-packages}

To download, clean, plot, and analyze the data, the statistical software R was used [@citeR]. Other packages such as tidyverse @tidyverse, janitor [@janitor], dplyr [@dplyr], ggplot2 [@ggplot2], lubridate [@lubridate], tinytex [@tinytex], zoo [@zoo], tidyr [@tidyr], knitr [@knitr], treemapify [@treemapify], and lintr [@lintr] aided in this process. In the data cleaning process, about 30 missing observations were found. These 30 observations were all under the variable of location type, demonstrating the report writer had a hard time describing the location of the incident. To deal with this, all the location was set to 'Unspecified' which can still be interpreted in the graphs. Some variables were transformed in the data to make graphs more interpretable, the [appendix](#appendix) documents how exactly other variables were changed.

## Observations {#sec-observations}

A sample observation of the cleaned data is shown in @tbl-1a and @tbl-1b This table plots the variables of interest for our analysis. It can be seen that most perpetrators are generally not prejudiced to more than one group. Also, due to the categorical nature of the data, this is the best 'summary statistic' that can be shown, although other displays of the data that show all observations are plotted in the @sec-data-analysis-and-plots section. This shows us that every variable of interest is documented rather nicely for analysis.

```{r}
#| echo: false
#| label: tbl-1a
#| page-break-before: auto

# Set the seed for reproducibility
set.seed(31459)

# Table 1: Display the first set of variables
random_sample_table_1 <- hate_crime_analysis_data %>%
  select(occurrence_date, location_type, hood_158,
         race_bias, ethnicity_bias) %>%
  sample_n(5)

# Display the first table
knitr::kable(random_sample_table_1,
             caption = "Random Sample of 5 Hate Crime Observations - Part 1",
             col.names = c("Occurence Date",
                           "Location Type",
                           "Neighbourhood Number",
                           "Race Bias",
                           "Ethnicity Bias"))
```

```{r}
#| label: tbl-1b
#| echo: false
#| page-break-before: auto
#Use the same seed to ensure these are the same observations.
set.seed(31459)

# Table 2: Display the second set of variables
random_sample_table_2 <- hate_crime_analysis_data %>%
  select(language_bias, religion_bias,
         sexual_orientation_bias, primary_offence,
         multiple_bias) %>%
  sample_n(5)  # The same random observations will be selected due to the seed

# Display the second table
knitr::kable(random_sample_table_2,
             caption = "Random Sample of 5 Hate Crime Observations - Part 2",
             col.names = c("Language Bias",
                           "Religion Bias",
                           "Sexual Orientation Bias",
                           "Primary Offence",
                           "Multiple Bias"))

```
## Data and Variable Selection: {#sec-data-and-variable-selection}

Only one other data set documenting these hate crimes exist, and come directly from miscellaneous police data. Since this data set is based off of the miscellaneous data (and is cleaner) the decision was made to use this specific data set. Certain variables were not explored due to being outside the scope of our paper: these include primary offence, and neighborhood. While the primary offence is an important variable of interest, plotting and analyzing it proves to be a challenge due to the low number of offences of each type. Also this paper tries to put an emphasis victims rather than the perpetrators of hate crime to get a better understanding of who is most vulnerable. Neighborhood analysis also proves troublesome to explore: since the data is only based on where the crime is committed, we can't infer too much. We don't know if either the victim or the perpetrator even lives in the area, so connecting it with other neighborhood variables such as income would leave out key parts of the story. For that reason, this paper only explores the group trends of the victims and general trends over time, rather than spatial analysis.

## Data Analysis and Plots {#sec-data-analysis-and-plots}
```{r}
#| label: fig-1
#| echo: false
#| fig.align: center
#| fig.width: 8
#| warning: false  # Hide one annoying error message
#| fig-pos: 'H'
#| fig-cap: Monthly Hate Crime Counts from 2018 to 2024,
#|           with annual moving average.
#| page-break-before: auto

# Extract year and month from the 'occurrence_date' column
hate_crime_analysis_data <- hate_crime_analysis_data %>%
  mutate(year_month = floor_date(occurrence_date, "month"))

# Group by year and month, then count observations
monthly_counts <- hate_crime_analysis_data %>%
  group_by(year_month) %>%
  summarise(count = n())

# Calculate a 12-month (yearly) moving average
monthly_counts <- monthly_counts %>%
  mutate(moving_avg = rollmean(count, 12, fill = NA,
                               align = "right"))

# Reshape the data for easier plotting of multiple lines
monthly_counts_long <- monthly_counts %>%
  pivot_longer(cols = c(count, moving_avg),
               names_to = "Line_Type",
               values_to = "Value")

# Plot with legend for monthly total and moving average.
# Lines: One black one blue, blue is moving average.
ggplot() +
  geom_line(data = monthly_counts_long,
            aes(x = year_month, y = Value,
                color = Line_Type, linetype = Line_Type),
            linewidth = 1) +  # Line plot for both trends
  geom_point(data = monthly_counts,
             aes(x = year_month, y = count),
             color = "black") +  # Points for monthly totals
  scale_color_manual(values = c("count" = "black",
                                "moving_avg" = "blue"),
                     labels = c("Monthly Total",
                                "12 Month Moving Average")) +
  scale_linetype_manual(values = c("count" = "solid",
                                   "moving_avg" = "dashed"),
                        labels = c("Monthly Total",
                                   "12 Month Moving Average")) +
  labs(
    title = "Number of Hate Crimes Per Month in Toronto",
    x = "Month",
    y = "Reported Hate Crimes",
    color = "Legend",
    linetype = "Legend"
  ) +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```
@fig-1 documents the evolution of hate crimes in Toronto over time. It plots the monthly quantity of hate crimes committed for all months from January 2018 to December 2023. It also plots the 12 month moving average to demonstrate the trend over time. The number of hate crimes go from averaging 12 a month in a given year in the beginning, to nearly 30 by the end of the data set. The catalyst for this upward trend seems to start during the early COVID-19 period in 2020, demonstrating that COVID-19 could be a possible explanatory factor this uptick. Interestingly, hate crimes still remained rather elevated post lock down, and in 2023 climbed to even higher levels in comparison to the COVID-19 period. May 2023 had the highest amount of hate crimes committed in Toronto since the start of this data, with 44 being committed. Hate crimes also exhibit a seasonal pattern, with December having the a relatively low amount of hate crimes in every single year.
```{r}
#| label: fig-2
#| fig-cap: Groups targeted by hate crime.
#| echo: false
#| fig-pos: 'H'
#| page-break-before: auto

# Step 1: Filter and reshape the data for non-multiple bias
long_data <- hate_crime_analysis_data %>%
  filter(multiple_bias == "NO") %>%  # Exclude entries with multiple bias
  select(occurrence_date, race_bias, ethnicity_bias, religion_bias,
         language_bias, age_bias, sexual_orientation_bias) %>%
  pivot_longer(cols = c(race_bias, ethnicity_bias, religion_bias,
                        language_bias, age_bias, sexual_orientation_bias),
               names_to = "bias_type", values_to = "bias_value") %>%
  filter(bias_value != "None" & bias_value != "NO")  # Exclude entries

# Step 2: Modify bias_type to represent the category
long_data <- long_data %>%
  mutate(bias_type = case_when(
    bias_type == "race_bias" ~ "Race Bias",
    bias_type == "ethnicity_bias" ~ "Ethnicity Bias",
    bias_type == "religion_bias" ~ "Religion Bias",
    bias_type == "language_bias" ~ "Language Bias",
    bias_type == "age_bias" ~ "Age Bias",
    bias_type == "sexual_orientation_bias" ~ "Sexual Orientation Bias",
    TRUE ~ bias_type  # Fallback case
  ))

# Step 3: Count hate crimes by year and bias type
hate_crime_counts <- long_data %>%
  mutate(year = year(occurrence_date)) %>%  # Extract year
  group_by(year, bias_type) %>%             # Group by year and bias type
  summarise(count = n(), .groups = "drop")  # Count entries

# Step 4: Count hate crimes where multiple_bias is TRUE
multiple_bias_counts <- hate_crime_analysis_data %>%
  filter(multiple_bias == "YES") %>%
  mutate(year = year(occurrence_date)) %>%
  group_by(year) %>%
  summarise(count = n(),
            bias_type = "Multiple Bias",
            .groups = "drop")  # Count entries

# Step 5: Combine the data for the final chart
final_counts <- bind_rows(hate_crime_counts, multiple_bias_counts)

# Step 6: Create the segmented bar chart
ggplot(final_counts, aes(x = factor(year), y = count, fill = bias_type)) +
  geom_bar(stat = "identity", position = "stack") +  # Use 'identity'
  labs(title = "Hate Crimes by Bias Type and Year",
       x = "Year",
       y = "Total Number of Hate Crimes",
       fill = "Bias Type") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +  # Set color palette
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis
```
@fig-2 plots the number of people targeted for each crime bias type. It can be seen that most hate crimes target only one specific characteristic of an individual, the most common of which are religion and race. The least common biases were age bias and language bias, where age bias didn't have a single observation and language only had one. Here we also see the same trend outlined in @fig-1, that is, hate crimes increased in 2020, and have remained at high levels since. In 2023 have seen even more increases in hate crime counts. This increase in 2023 is generally fueled by increases in hate crime against religion, where as the increases in 2020 and 2021 are generally fueled by race. This suggests that the reasons for the increases in the hate crimes during those periods likely different.

```{r}
#| label: fig-3
#| fig-cap: Proportion of Total Hate Crimes by Location Type
#| echo: false
#| fig-width: 8
#| fig-pos: 'H'
#| page-break-before: auto

# Aggregate the data by location_type
location_type_counts <- hate_crime_analysis_data %>%
  group_by(location_type) %>%
  summarize(total_crimes = n())

# Create the pie chart
ggplot(location_type_counts, aes(x = "", y = total_crimes,
                                 fill = location_type)) +
  geom_bar(stat = "identity", width = 1) +  # Bar chart to be pie chart
  coord_polar(theta = "y") +  # Convert to pie chart
  labs(title = "Breakdown of Hate Crimes by Location Type",
       fill = "Location Type") +
  scale_fill_brewer(palette = "Set3") +  # RColorBrewer color palette
  theme_void() +  # Clean pie chart
  theme(plot.title = element_text(hjust = 0.5))  # Center title

```
@fig-3 Plots the relative proportions of all locations types for hate crimes. There is a large variety of the kinds of locations where these events take place. This could perhaps suggest that hate crimes are spur of the moment attacks rather than premeditated ones, as the locations do not seem to happen in any one fixed location. The most common locations are streets/roads/highways, residential areas, and educational institutions. The [appendix](#appendix) plots in a segmented bar chart to show dynamics over time, however, the results don't vary much over time.


```{r}
#| label: fig-4
#| fig-cap: Yearly Breakdown of Race-Motivated Hate Crimes
#| warning: false
#| echo: false
#| fig-pos: 'H'
#| page-break-before: auto

ggplot(hate_crime_analysis_data %>%
         filter(race_bias != "None"),
       aes(x = factor(year(occurrence_date)),
           fill = race_bias)) +
  geom_bar(position = "dodge") +  # Stack bars to show total counts by year
  labs(title = "Hate Crimes by Race and Year",
       x = "Year",
       y = "Total Hate Crimes",
       fill = "Race Bias") +
  scale_fill_brewer(palette = "Set1") +  # RColorBrewer color palette
  theme_minimal()
```
@fig-4 provides the year breakdown of race-motivated hate crimes. As seen, Black people are clearly the most targeted group for race-motivated hate crimes in Toronto. They made up the largest proportion of hate crime victims in every single year, and were almost the only victims of race targeted hate crimes in 2018 and 2019. Another significant subgroup were East/Southeast Asians. They are particularly notable in 2021 onward, suggesting that COVID based rhetoric may have played a role in East/Southeast Asians becoming victims of hate crimes.


```{r}
#| label: fig-5
#| fig-cap: Yearly Breakdown of Religion-Motivated Hate Crimes
#| echo: false
#| fig-pos: 'H'
#| warning: false
#| page-break-before: auto

# Load the palette from RColorBrewer
palette_colors <- RColorBrewer::brewer.pal(n = 10, name = "Set2")

# Modify "Other Christian Religion" color while keeping the rest of the palette
custom_colors <- c(
  palette_colors[1],  # First color
  palette_colors[2],  # Second color
  palette_colors[3],  # Third color
  palette_colors[4],
  palette_colors[5],
  palette_colors[6],
  palette_colors[7],
  palette_colors[8],
  palette_colors[9],
  "Other Christian Religion" = "#0072B2",  # Custom color
  palette_colors[10]
)

ggplot(hate_crime_analysis_data %>%
         filter(religion_bias != "None"),
       aes(x = factor(year(occurrence_date)),
           fill = religion_bias)) +
  geom_bar(position = "stack") +  # Stack bars to show total counts by year
  labs(title = "Hate Crimes by Religion and Year",
       x = "Year",
       y = "Total Hate Crimes",
       fill = "Religion Bias") +
  theme_minimal()
```
@fig-5 provides the year breakdown of religion motivated hate crimes. Consistently throughout 2018 to 2023, Jewish Torontonians have faced the largest number of religion-motivated hate crimes, taking up at least 80 percent of the crimes. The only other group that faced significant hate crimes were Muslims, who faced most of the remaining crimes. Up until 2022, the number of religion motivated hate crimes stayed largely the same, but 2023 saw an approximate 75% upswing in religion based hate crimes. This paper is not unfortunately able to point to the causal effects for these upswings, which will be discussed further in @sec-discussion-and-limitations.

# Discussion and Limitations {#sec-discussion-and-limitations}

The data provides information on the nature of hate crimes in Toronto. @fig-1 tells us that hate crimes have grown rapidly over time and particularly exploded in 2020 and 2023. @fig-2 tells us that the most common prejudices hate crime perpetrators have is against religion and race, and that religion based hate crime has grown over time. @fig-3 tells us that hate crimes can happen almost anywhere, and there isn't one universal place where they happen. @fig-4 and @fig-5 tell us that in the race and religion subgroups, it is Black and Jewish individuals who face the majority of the hate crimes. Overall, these trends tell us that hate crime is a growing issue in Toronto and policy needs to try and address it before if Toronto wants to maintain is multicultural and accepting status. The fact that the victim group can vary so largely depending on the time period in even a short 5 year sample like this means that policy makers have to play close attention to rhetoric in times of tension. Almost no kind of hate crime has decreased in Toronto since its explosion in 2020 (perhaps except Asian targeting hate crime), and this suggests that not enough is being done to tackle it.

While the paper does observe changes in the types of hate crimes and the quantity of hate crimes experienced over time, what it lacks is some causal analysis. I cannot determine what are the reasons for the the large upswings in Toronto in specific in 2020, 2021, and 2023, or why the 'average victim' changes over time. While due to the timing one may be able to guess that COVID played a role, nothing can be said for sure until a causal research design is conducted for Toronto in specific. These could be broken up into a few different studies – what are the determinants of the rise in East/Southeast Asian in 2021, but it's subsequent dissipation post 2021. Other areas of research could focus on why the breakdowns of hate crimes are the way they are, even if there were not huge changes over time. Why is it that Jewish and Black Torontonians take up the bulk of the hate crimes, and are there successful ways to stop the growth of attacks on these individuals? 

Another drawback of my paper is the lack of information I have about the perpetrators of the hate crime. This was on purpose: the victim data was  more readily available and interpretable, whereas the perpetrator data was more limited and had more caveats. The only data available for the perpetrator (other than their bias type) was where the crime committed. Even with this, I could not match the location to where a hate crime offender lived to get more background characteristics. This limited me to focusing on just the who a 'typical victim' was, rather than a 'typical offender'.

There are huge gaps in the literature to explore the mechanisms that drive hate crime variance over time and groups. Exploring these mechanisms could ultimately give us ways that hate crime could be reduced by policy in the future.

# Appendix {#appendix .appendix}

Some variables were altered in order to create more informative graphs, the appendix documents exactly how these variables were changed.

## Race Share {.appendix}

In order to reduce the number of categories, any biracial hate crime victims were identified by their primary race. A previous observation in the data would be "Black, Hispanic", the changes conducted in our data cleaning section would change it just Black. This prevented us from having upwards of 15 groups which would be hard to plot categorically.

## Location Type {.appendix}

Location type was changed for a similar reason: apartments and houses were combined into a singular category: residential. This would also be more informative since they serve the same purpose. Previously, there were two government building types: "Government Building", and "Government Building (Courthouse, Museum, Parliament etc, this edit combined them into a "Government Building".

## Plot of Location Types In Bar Chart Form {.appendix}

In @fig-6, we don't see much variation in the locations of hate crimes over time. The relative proportion stays rather constant, suggesting even the increase in hate crimes didn't change much where they happened. This further supports the idea of hate crimes being spur of the moment attacks.

```{r}
#| label: fig-6
#| fig-cap: Yearly Breakdown of Hate Crimes by Location Type
#| echo: false
#| fig-width: 10
#| fig-pos: 'H'

ggplot(hate_crime_analysis_data,
       aes(x = factor(year(occurrence_date)),
           fill = location_type)) +
  geom_bar(position = "fill") +  # Group the bars by location type
  labs(title = "Hate Crimes by Location Type and Year",
       x = "Year",
       y = "Total Hate Crimes",
       fill = "Location Type") +
  scale_fill_brewer(palette = "Set3") +  # Choose a color palette
  theme_minimal()
```

\newpage

\newpage

# References
