---
title: "Toronto's Uptick in Hate Crime: Time and Group Trends"
subtitle: "My subtitle if needed"
author: 
  - Justin Klip
thanks: "Code and data are available at: https://github.com/justinklip/hate_crime_toronto_paper."
date: today
date-format: long
abstract: "Using data from Open Data Toronto, this paper the tracks the evolution of hate crime in Toronto from 2018 to the end of 2023. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(dplyr)
library(ggplot2)
library(lubridate)
library(palmerpenguins)
library(tinytex)

#Data Acquisition
hate_crime_analysis_data <- read_csv(here::here("data/analysis_data/hate_crime_analysis_data.csv"))
```

# Introduction

In 1988 the Canadian Parliament passed the Canadian Multiculturalism Act, an act that served to recognize and preserve Canada's multicultural heritage (Government of Canada 2024). Toronto serves as a shining example of Canada's multiculturalism, with over half of the population of Toronto being foreign-born (McNeil). Yet, despite Toronto's reputation as an international city with large widespread cultural heterogeneity, hate crimes have seen an uptick in recent times. Hate crimes in Toronto have increased nearly 55% in 2024 in comparison to 2023. A hate crime is defined as a "criminal act done by a person who is motivated by an extreme bias or hatred towards a particular social group" (Government of Canada 2023). Acts like these are fundamentally at odds with the sentiment expressed by the Canadian Multiculturalism Act and raise questions about Toronto's history with hate crimes and its multicultural status. In this paper, I explore the dynamics of Toronto's Hate Crimes from 2018 to 2023, before the current uptick. In particular, I describe how the types of hate crimes, the types of victims, and the quantity of victims have changed over time.

The rest of the paper is as follows: I first describe the data set, how it is used, and how it was generated in @data. Then in the @results section, I plot the data to highlight COVID's role in elevating hate crime counts and show how the 'average victim' has changed over time. In the @discussion section, I interpret these results and what they mean in terms of policy implications. Then, in the @limitations-and-next-steps section, I explain what this paper lacks and what future research in this field could focus on.

# Data {#sec-data}

## Measurement

To run my analysis, data was collected from Open Data Toronto's Open Data Portal, through the R library @opendatatoronto. The data source used is their Hate Crimes Open Data data set. This data of about 1400 observations has every crime that has been classified as a 'hate crime' by Toronto Police from the beginning of 2018 to the end of 2023. The data is generated as follows: a crime is committed in Toronto, and generally within a month that crime is reported by the victim. If a hate-motivated offence is suspected, the investigation is then led by a divisional investigator of the Hate Crime Unit (HCU) or by the HCU itself (Toronto Police Service). If hate-motivated offences are found, then this case will be added next year to the data set. Details of the crime are documented – including what social group they were biased against – these bias variables include ethnicity, race, gender, age, language and sexual orientation. Date and time (of both the crime and the reporting of it), offence committed, neighborhood, and place (such as apartments, malls, parks) were also recorded. Every unique case number was also attributed to each observation in the data set.

## Cleaning and Packages

To clean and analyze the data, the statistical software R was used [@citeR].

```{r}
#| label: fig-monthlyhatecrimes
#| fig-cap: Number of reported hate crimes from month to month in Toronto
#| echo: false


# Extract year and month from the 'occurrence_date' column
hate_crime_analysis_data <- hate_crime_analysis_data %>%
  mutate(year_month = floor_date(occurrence_date, "month"))

# Group by year and month, then count observations
monthly_counts <- hate_crime_analysis_data %>%
  group_by(year_month) %>%
  summarise(count = n())

# Plot the number of observations per month, with x-axis labels every quarter
ggplot(monthly_counts, aes(x = year_month, y = count)) +
  geom_line() +   # Use line plot to show trends over time
  geom_point() +  # Add points for better visualization
  labs(
    title = "Number of Hate Crimes Per Month in Toronto",
    x = "Month",
    y = "Reported Hate Crimes",
    caption = "Figure 1: Monthly Hate Crime Data from 2018 to 2024"
  ) +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.caption = element_text(hjust = 0.5)  # Center the caption
  )

```

@fig-monthlyhatecrimes

```{r}

```

Talk way more about it.

# Results

```{r}

```

```{r}

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Limitations

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

```{r}

```

## Diagnostics

```{r}

```

\newpage

# References
